---
title: "AD Annotation Schema Testing"
author: "Nicole Kauer"
date: "4/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Requirements & Helper Functions

### Schema Registration

Schema are registered to an organization. To register a schema, you must have the appropriate permissions on the organization ACL.

### Running This Code

This requires `reticulate` to be installed and able to import the Synapse python client. If you do not have `reticulate` set up, then the following snippet _usually_ works.

```{r reticulate, eval=FALSE}
install.packages('reticulate')
reticulate::install_miniconda()
reticulate::py_install(c("numpy", "pandas", "synapseclient"), pip = TRUE)
```

The following packages are used. Please install them, if needed.

```{r packages, eval=FALSE}
install.packages(c("purrr", "glue", "reticulate", "here"))
```

Import libraries.
```{r libraries, message=FALSE, eval=FALSE}
library("purrr")
library("glue")
library("reticulate")
library("here")
```

These functions are for assisting with registration and access of the schema. Note that most of these functions are in the synapseAnnotations repository, but adapted for use with `reticulate`.


```{r helpers, message=FALSE}
# Functions to register JSON schema file ---------------------------------------

#' Create schema request body
#'
#' @param file Path to JSON Schema file
#' @param dryRun Should the dryRun field be set to `true`? Defaults to `FALSE`.
#' @return String containing the JSON request body
create_body <- function(file, dryRun = FALSE) {
  stopifnot(dryRun %in% c(TRUE, FALSE))
  paste(
    '{"concreteType": "org.sagebionetworks.repo.model.schema.CreateSchemaRequest", "schema": ',
    paste(readLines(file), collapse = ""),
    ', "dryRun": ',
    ifelse(dryRun, 'true', 'false'),
    '}'
  )
}

#' Register schema to Synapse
#'
#' Given a path to a file containing a JSON schema, registers that schema on
#' Synapse.
#'
#' @param syn Synapse object
#' @param file Path to JSON Schema file
#' @param dryRun Should the dryRun field be set to `true`? Defaults to `FALSE`.
#' @return Token object containing the async token used to monitor the job
register_schema <- function(syn, file, dryRun = FALSE) {
  syn$restPOST(
    uri = "/schema/type/create/async/start",
    body = create_body(file, dryRun = dryRun)
  )
}

#' Get status of registered schema
#'
#' @param syn Synapse object
#' @param token Async token
#' @return Response object containing schema
get_registration_status <- function(syn, token) {
  processing <- TRUE
  ## Check results of registering schema, retrying if the async job hasn't
  ## completed yet
  while (processing) {
    result <- syn$restGET(
      uri = glue("/schema/type/create/async/get/{token}")
    )
    ## synapser doesn't return the status codes unfortunately, so we check the
    ## response object to determine what to do. If it contains "jobState",
    ## that's part of the AsynchronousJobStatus, i.e. the async job hasn't
    ## completed yet.
    if (!"jobState" %in% names(result)) {
      processing <- FALSE
    }
  }
  result
}

#' Register schema and report on its status
#'
#' Registers a schema and then monitors the asynchronous job until it completes.
#'
#' @param syn Synapse object
#' @param file Path to JSON Schema file
#' @param dryRun Should the dryRun field be set to `true`? Defaults to `FALSE`.
#' @return `TRUE` if schema was successfully registered; `FALSE` otherwise.
register_and_report <- function(syn, file, dryRun = FALSE) {
  message(glue("Registering {file} with dryRun = {dryRun}"))
  ## register_schema() can fail if e.g. schemas contain non-ASCII characters --
  ## in this case we catch that error and provide the message.
  token <- try(register_schema(syn, file, dryRun = dryRun), silent = TRUE)
  if (inherits(token, "try-error")) {
    message(token)
    return(FALSE)
  }
  ## Note that this is written for how synapser works.
  ## It should be updated for reticulate, but it currently works as is.
  ## Reasoning for synapser below:
  ##   If we get a 400 status code, synapser will throw an error -- we want to
  ##   catch that and provide the message but not raise an actual R error.
  status <- try(get_registration_status(syn, token$token), silent = TRUE)
  if (inherits(status, "try-error")) {
    message(status)
    FALSE
  } else {
    TRUE
  }
}
```

## Usage

Import the `synapseclient` and log into Synapse.

```{r synapse, eval=FALSE}
## Log in to synapse
synapse <- import("synapseclient")
syn <- synapse$Synapse()
syn$login()
```

### Registering Schema

First find the schema files.

```{r files, eval=FALSE}
files <- list.files(
  here("schema"),
  pattern = ".json$",
  full.names = TRUE
)
```

For schemas that use referencing, there is an order to how schemas are registered. All referenced schema must be registered before they can be referenced. For the schema in this project, I've numbered the schema to indicate at what 'level' they are: a higher number means the sooner it must be registered. For example, the main schema is at level 01 and references both schema at level 02. The schema at level 02 must be registered before the main schema, at level 01, can be registered. The nice thing about labeling the level at the front of the file name is that the files are already ordered from level 01 to 05. To register these in the correct order, all that's needed is to reverse the order.

```{r reverse_files, eval=FALSE}
files <- rev(files)
```

Some of these might already be registered. If these were versioned, it would be impossible to register them a second time. In the case of versioned schema, do a dry-run first to check which ones do not need to be registered. The following example snippet will return a boolean list that can be used to filter out the schemas that already registered in Synapse (i.e. boolean value is `FALSE`). 

**Note:** Any `FALSE` returns from the dry-run could be from the schema already existing in Synapse or from the schema being altogether invalid. Here, we are assuming the schemas are valid and that any `FALSE` is purely due to not being registered, yet. Again, if a schema is not versioned and the file is valid, this will return `TRUE`, even if the schema is identical to latest registered version.

```{r dryrun, message=FALSE, eval=FALSE}
should_register <- map_lgl(files, register_and_report, syn = syn, dryRun = TRUE)
```

Check to see if any need registering and, if so, filter the list of files.
```{r register, eval=FALSE}
if(any(should_register)) {
  ## Only keep `TRUE` files
  files <- files[should_register]
} else {
  files <- NA
}
```

Register files by looping through all the files that should be registered. Output will be `TRUE` for each file that was registered.

```{r registration, message=FALSE, eval=FALSE}
if(any(should_register)) {
  ## Register files
  map_lgl(files, register_and_report, syn = syn) 
}
```
### Binding Schema

Any schema that is registered in Synapse can be bound to a project, directory, or other entity in Synapse.

```{r bind, eval=FALSE}
bind_schema <- function(syn, entity_id, schema_id) {
  ## Build strings outside of rest call for reasons
  uri <- glue::glue("/entity/{entity_id}/schema/binding")
  body <- glue::glue("{{entityId: \"{entity_id}\", schema$id: \"{schema_id}\"}}")
  syn$restPUT(
    uri = uri,
    body = body
  )
}
```

Use helper function above to bind a schema to an entity. In this example, I binding the nkauer-ad.main schema to the Studies directory in my test project.

```{r binding, eval=FALSE}
desired_entity_id <- "syn25421165"
desired_schema_id <- "nkauer-ad.main"

bind_schema(syn, desired_entity_id, desired_schema_id)
```
### Get Validation Schema

Synapse needs to create a validation schema based on the registered schema. This is especially important for schema with references to other schema; Synapse needs to follow the references and gather the needed information. You can pull down the validation schema that Synapse creates.

```{r validation_schema}
get_validation_schema <- function(syn, schema_id) {
  ## Start the job to compile the validation schema
  token <- syn$restPOST(
    uri = "/schema/type/validation/async/start",
    body = glue::glue("{{$id: \"{schema_id}\"}}")
  )
  ## Should wait until finished
  schema <- get_schema_status(syn, token)
  schema
}
## Have to keep checking on status
get_schema_status <- function(syn, token) {
  processing <- TRUE
  ## Check results of registering schema, retrying if the async job hasn't
  ## completed yet
  while (processing) {
    result <- syn$restGET(
      uri = glue::glue("/schema/type/validation/async/get/{token}")
    )
    ## synapser doesn't return the status codes unfortunately, so we check the
    ## response object to determine what to do. If it contains "jobState",
    ## that's part of the AsynchronousJobStatus, i.e. the async job hasn't
    ## completed yet.
    if (!"jobState" %in% names(result)) {
      processing <- FALSE
    }
  }
  result
}
```

Here is an example that grabs the nkauer-ad.main validation schema.

```{r example_schema, eval=FALSE}
desired_schema_id <- "nkauer-ad.main"
validation_schema <- get_validation_schema(syn, desired_schema_id)
```

This is a large schema. Can turn it into prettified json with jsonlite.

```{r pretty_json, eval=FALSE}
json_string <- jsonlite::prettify(jsonlite::toJSON(validation_schema))

## Recommend writing to a file versus looking at the string in R
write(json_string, file = "nkauer-ad.main_dereferenced.json")
```

### Validate Annotations Using Schema

The title of this section is a trick because Synapse is already working on validating every file against the schema. You don't have to do anything other than wait patiently.

### Get Validation Results

Grabbing the validation results is simple.

```{r validation}
get_validation_results <- function(syn, entity_id) {
  syn$restGET(
    uri = glue::glue("/entity/{entity_id}/schema/validation")
  )
}

## Check if results say the file is valid or invalid
## Overly simplistic function...
is_valid_result <- function(validation_result) {
  validation_result$isValid
}
```

There is a table with all files that are in my 'Studies' directory, which are bound to the nkauer-ad.main schema. Loop through all of them to get their annotation validation status.

```{r results, eval=FALSE}
## Grab all synIDs that have a schema bound
## I have these in a fileview for ease
fileview <- "syn25421275"
all_files <- syn$tableQuery(
  glue::glue("SELECT id FROM {fileview}")
)$asDataFrame()$id

all_results <- purrr::map(all_files, ~ get_validation_results(syn, .))
is_valid <- purrr::map_lgl(all_results, ~ is_valid_result(.))

## Check if all or any are valid
any(is_valid)
```

Parsing the returned validation results for invalid annotations is left as an exercise...
